---
title: "Reproducible Research: Peer Assessment 1"
author: "Morakinyo"
date: "9/16/2023"
output: 
  html_document:
    keep_md: true
---

```{r, echo=TRUE}
library(ggplot2)
library(dplyr)
library(lubridate)

options(dplyr.summarise.inform = FALSE)
```
## Loading and preprocessing the data

```{r, echo=TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = paste0(getwd(), '/repdata%2Fdata%2Factivity.zip'))
unzip("repdata%2Fdata%2Factivity.zip")

activity <- read.csv('activity.csv')
head(activity)
```


Doing some exploratory analysis

```{r, echo=TRUE}
dim(activity)
summary(activity)
str(activity)
```


## What is mean total number of steps taken per day?

### The Total number of steps taken per day

```{r, echo=TRUE}
totalsteps <- aggregate(steps~date, activity,sum)
totalsteps
```

### A Histogram of the total number of steps taken each day
```{r, echo=TRUE}
ggplot(totalsteps, aes(x = steps)) + geom_histogram(binwidth = 2000) + xlab("Total number of steps per day") + ylab("Frequency")
```

```{r, echo=TRUE}
mean = mean(totalsteps$steps, na.rm = TRUE)
median = median(totalsteps$steps, na.rm = TRUE)
mean
median
```

The mean and median are as follows respectively:

    1. 10766.19
    2. 10765

## What is the average daily activity pattern?

### Time series plot of the average number of steps taken

```{r, echo=TRUE}
rm(mean)
avsteps <- aggregate(steps~interval, activity, mean)

ggplot(data=avsteps, aes(x=interval, y=steps)) +
  geom_line() +
  ggtitle("Time Series Plot of the Average Number of Steps") +
  xlab("Interval") + 
  ylab("Average number of steps taken")

  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
highest_steps =avsteps %>% filter(steps == max(steps))
highest_steps
```

## Imputing missing values

```{r, echo=TRUE}
missing <- sum(!complete.cases(activity))
```

```{r, echo=TRUE}
data_impute <- activity %>%
  mutate(
    steps = case_when(
      is.na(steps) ~ avsteps$steps[match(activity$interval, avsteps$interval)],      
      TRUE ~ as.numeric(steps)
    ))
```


## A Histogram of the total number of steps taken each day after missing values are imputed

```{r, echo=TRUE}
total_steps <- data_impute %>% group_by(date) %>% summarise(daily_steps = sum(steps))

ggplot(total_steps, aes(daily_steps)) + 
  geom_histogram(binwidth = 2000) + 
  xlab("Total number of steps per day") + 
  ylab("Frequency")
```

## Are there differences in activity patterns between weekdays and weekends?

### Creating a new factor variable in the dataset with two levels - "Weekday" and "Weekend".

```{r, echo=TRUE}
day_of_week <- data_impute %>%
  mutate(
    date = ymd(date),
    weekday_or_weekend = case_when(wday(date) %in% 2:6 ~ "Weekday",
                                   wday(date) %in% c(1,7) ~ "Weekend")
  ) %>% select(-date) %>%
  group_by(interval, weekday_or_weekend) %>%
  summarise(
    steps = mean(steps)
  )
```


### Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends.
```{r}
ggplot(day_of_week, aes(interval, steps)) + 
  geom_line() + 
  facet_wrap(~weekday_or_weekend, nrow = 2) +
  xlab("Interval") + 
  ylab("No. of Steps")
```